---
title: "R Practice — Bias in Self-Reported Turnout"
author: "Student Name"
date: "Quantitative Methods for Humanities. PPE 2nd Year - 2025-2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

> **Important**  
> 1) Fill in `student_id` below.  
> 2) Do **not** rename chunk labels (e.g., `ex1`, `ex2`, ...).  
> 3) Your file must **Knit** (Compile) to HTML from a **clean session** (no manual steps).

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(stringsAsFactors = FALSE)

# Include here all libraries you will later need

# Fill this in (used for reproducibility / individualized data)
student_id <- "REPLACE_WITH_YOUR_ID" # It should be a number

# If your instructor provides individualized datasets using a seed, this is the standard pattern:
set.seed(student_id)

# --- DO NOT CHANGE OR REMOVE THIS (IT'LL RESULT IN A GRADE OF ZERO)
if (!is.numeric(student_id) || length(student_id) != 1) stop("student_id must be a single number.")
```

---

# Introduction

Surveys are frequently used to measure political behavior such as voter turnout, but some researchers are concerned about the accuracy of self-reports. In particular, they worry about possible social desirability bias where, in postelection surveys, respondents who did not vote in an election lie about not having voted because they may feel that they should have voted. Is such a bias present in the American National Election Studies (ANES)? ANES is a nationwide survey that has been conducted for every election since 1948. ANES is based on face-to-face interviews with a nationally representative sample of adults. 

# Data 

The following table displays the names and descriptions of variables in the  `turnout.csv` data file.

| Variable      | Description                                             |
| ------------- | ------------------------------------------------------- |
| `year`        | election year                                           |
| `ANES`        | ANES estimated turnout rate                             |
| `VEP`         | voting eligible population (in thousands)               |
| `VAP`         | voting age population (in thousands)                    |
| `total`       | total ballots cast for highest office (in thousands)    |
| `felons`      | total ineligible felons (in thousands)                  |
| `noncit`      | total noncitizens (in thousands)                        |
| `overseas`    | total eligible overseas voters (in thousands)           |
| `osvoters`    | total ballots counted by overseas voters (in thousands) |


## Exercise 1 - Data summary 

**Tasks/Questions:**

1. Load the data into R and save it into a variable called `turnout` (place the dataset in the same folder as this .Rmd).
2. Check the dimensions. 
3. Obtain a summary of the dataset. 
4. Report: number of observations and range of years covered.

```{r ex1}
# TODO: add your code to answer here

```

```{r data-randomization, include=FALSE}
# --- Data randomization : DO NOT CHANGE OR REMOVE THIS (IT'LL RESULT IN A GRADE OF ZERO)
if (!exists("turnout") || !is.data.frame(turnout)) stop("You must load the dataset into a data frame called `turnout` before running the randomization.")
count_cols <- c("ANES","VEP","VAP","total","felons","noncit","overseas","osvoters")
row_jitter <- runif(nrow(turnout), 0.995, 1.005)  # ±0.5% per row (same factor for all counts)
turnout[count_cols] <- lapply(turnout[count_cols], function(x) round(pmax(0, x * row_jitter), 1))
turnout$fingerprint <- round(sum(turnout$total) + 1000*mean(turnout$ANES), 3)
turnout$fingerprint
```

**Takeaways:**

* Add a few brief takeaways (no more than 3) based on your output/observations
* Include metrics from your analysis to backup your statements  (better if in-line code is used, rather than hard-coding numbers)

**Interpretation:** Add a brief interpretation

---

# Analysis

## Exercise 2 — Turnout rate calculation (VAP and VEP)

**Tasks/Questions:**

1. Compute turnout based on **VAP**, adding eligible overseas voters to the denominator (because `VAP` excludes them). Store it in the dataset as a variable called `vap_turnout`.
2. Compute turnout based on **VEP**. Store it in the dataset as a variable called `vep_turnout`.
3. Describe the difference you observe.

```{r ex2}
# TODO: add your code to answer here
```

**Takeaways:**

* Add a few brief takeaways (no more than 3) based on your output/observations
* Include metrics from your analysis to backup your statements  (better if in-line code is used, rather than hard-coding numbers)

**Interpretation:** Add a brief interpretation

---

## Exercise 3 — Compare VAP vs ANES, and VEP vs ANES

**Tasks/Questions:** 

1. Compute the differences between:
  - VAP and ANES estimates of turnout rates (store it in the dataset as a variable called `diff_vap_anes`).
  - VEP and ANES estimates of turnout rates (store it in the dataset as a variable called `diff_vep_anes`).
2. Report:
  - Average difference,
  - Range of the differences.

```{r ex3}
# TODO: add your code to answer here
```

**Takeaways:**

* Add a few brief takeaways (no more than 3) based on your output/observations
* Include metrics from your analysis to backup your statements  (better if in-line code is used, rather than hard-coding numbers)

**Interpretation:** Add a brief interpretation

---

## Exercise 4 — Compare across election types (presidential vs midterm)

**Tasks/Questions:** 

1. Create the variable `turnout$election_type` to classify elections year into `presidential` (multiple of 4) and `midterm` (non-presidential).
2. Compare **VEP turnout** vs **ANES** separately for:
  - presidential elections,
  - midterm elections.

Note: the dataset excludes **2006**.

```{r ex4}
# TODO: add your code to answer here
```

**Takeaways:**

* Add a few brief takeaways (no more than 3) based on your output/observations
* Include metrics from your analysis to backup your statements  (better if in-line code is used, rather than hard-coding numbers)

**Interpretation:** Add a brief interpretation

---

## Exercise 5 — Temporal bias trends (two periods)

**Tasks/Questions:** 

1. Has the bias of ANES increased over time? (use `turnout$diff_vep_anes` created in Exercise 3 and `turnout$diff_vep_anes` computed in Exercise 4).
2. Is the time-evolution pattern the same for presidential vs midterm years? 
3. Use the function `boxplot()` to visualize the differences.

```{r ex5}
# TODO: add your code to answer here
```

---

## Exercise 6 — Adjusted VAP turnout rate (2008)

**Tasks/Questions:** Do the following just for the year 2008:

1. Adjust VAP by subtracting **felons** and **noncitizens**. Store it in a variable called `adj_vap`.
2. Adjust turnout-rate numerator by subtracting **overseas ballots** (`osvoters`) from `total`. Store it in a variable called `adj_total`.
3. Compute the adjusted VAP turnout rate. Store it in `adj_vap_turnout_2008`.
4. Compare adjusted VAP turnout to: unadjusted VAP turnout, VEP turnout, and ANES turnout (in 2008).  
5. Briefly discuss.

```{r ex6}
# TODO: add your code to answer here
```

**Takeaways:**

* Add a few brief takeaways (no more than 3) based on your output/observations
* Include metrics from your analysis to backup your statements  (better if in-line code is used, rather than hard-coding numbers)

**Interpretation:** Add a brief interpretation

---

# Conclussions

Add quick conclusions

# Submission checklist
- ✅ `your_file.Rmd`  
- ✅ `your_file.html` (compiled output)


