---
title: "My First Markdown"
author: "Your Name"
output: html_document
---

# This is a Header (Chapter)

## This is a smaller header (Section)

### This is an even smaller header (Subsection)

#### You get the point

You are creating your first Rmarkdown document. Congrats!

# Exercise 4
Perform assignments:
```{r}
a <- 3 * (4 + 9)
b <- c(1, 4, 1999, 2011)
d <- 2 * pi
e <- a^2 + sqrt(d)
```

# Exercise 6
Create and analyze a vector:
```{r}
year <- seq(2010, 1950, by = -10)
print(year)
length(year)
min(year)
max(year)
range(year)
mean(year)
sum(year)
```

# Exercise 7
Import and analyze data:
```{r}
UNpop <- read.csv("UNpop.csv")
class(UNpop)
names(UNpop)
nrow(UNpop)
ncol(UNpop)
dim(UNpop)
summary(UNpop)
year_column <- UNpop$year
year_column
UNpop[4:5, ]
even.world.pop <- UNpop$world.pop[seq(2, nrow(UNpop), by = 2)]
even.world.pop
```

# Note
Use `{r echo=FALSE}` to suppress code output where necessary.

# Compile
Compile using the Knit button or the shortcut Ctrl + Shift + K to compile.